An event-sourced and decentralized database with IPLD and materialized views

### 2019-09-23 draft.tex

The use of well-established design patterns from the CQRS+ES (and related) literature means the Threads protocol can provide the foundation for many higher-level concepts and technologies. For instance, the public application programming interface (API) for a Thread provides several features that you would expect when operating on a dataset or table within a database. Indeed, each Thread is defined by a unique ID (its Identity), and provides facilities for access control and permissions, networking, sharing, invites, and more.

Like many CQRS-based systems, the ``public'' API for Threads revolves around the concept of Views, which are used to optimize queries on the local event store. A key feature of Views is they are completely disposable, because they can be entirely rebuilt from the source event store\footnote{Though see section \ref{sec:SnapshotsCompaction} for a discussion of Snapshots, which may be used to speed up Views.}. In Threads, Views are updated by subscribing to specific Event(s), and applying a Reducer function that is used to update the View state (which may be \emph{materialized} to local persistent storage). This simple subscription system provides a flexible framework for building complex View-driven application logic.

\subsubsection{Default Views}\label{sec:DefaultViews}

The core Textile Threads interfaces are also exposed via interfaces compatible with \emph{existing} datastores and/or communications systems.
These simple, public-facing APIs will be comfortable to application developers looking to leverage highly-distributed systems that connect user-controlled data, with minimal configuration and maximum interoperability.
In the remainder of this section, we highlight a number of planned database abstractions to provide further utility to developers interested in building on Textile's Threads protocol.

\paragraph{Databases}

A key/value store built on top of Threads can map database \texttt{Put} operations into internal \texttt{KeyValueUpdated} events. These Events would then operate on a map-like view model, via an internal ``projection'' (reducer function). The key/value view can be persisted or kept in-memory, and would be directly queried via a corresponding \texttt{Get} operation. Because Textile uses ES patterns, a \texttt{Del} operation would simply lead to an internal \texttt{KeyValueDeleted} event, marking the key/value pair for removal from the key/value materialized view (Example \ref{ex:KVStore}).

\begin{example}
\begin{minipage}{.45\textwidth}
\begin{lstlisting}
type TextileKVStore interface {
  Put(key string, value interface{}) error
  Get(key string) (interface{}, error)
  Del(key string) error
}
\end{lstlisting}
\end{minipage}
\caption{The Key-Value store interface.}
 \label{ex:KVStore}
\end{example}

Similar abstractions could (and will) be used to implement additional database types and functions. The key here is to create simple, higher-level interfaces on top of Textile Threads that simplifies dealing with events and data, but exposes the underlying event sourced data for customizing as needed. Developers should not have to learn a whole new set of terms and tools to take advantage of Textile's Threads capabilities. Other database abstractions include a no-sql style document store for storing and indexing arbitrary structs and/or JSON documents (Example \ref{ex:DocStore}).

>>>>

The use of well-established design patterns from the CQRS+ES (and related) literature means the Threads protocol can provide the foundation for many higher-level concepts and technologies. Like many CQRS-based systems, the ``public'' API for Threads revolves around the concept of Views, which are used to optimize queries on the local event store. A key feature of Views is they are completely disposable, because they can be entirely rebuilt from the source event store\footnote{Though see section \ref{sec:SnapshotsCompaction} for a discussion of Snapshots, which may be used to speed up Views.}. In Threads, Views are updated by subscribing to specific Event(s), and applying a Reducer function that is used to update the View state (which may be \emph{materialized} to local persistent storage). This simple subscription system provides a flexible framework for building complex View-driven application logic.

\subsubsection{Databse Views}\label{sec:DefaultViews}

The public API for a Thread already provides several features that you would expect when operating on a dataset or table within a database. Indeed, each Thread is defined by a unique ID (its Identity), and provides facilities for access control and permissions, networking, sharing, invites, and more. As such, it is not difficult to imagine a simplified API in which Threads are exposed via interfaces compatible with \emph{existing} datastores and/or communications systems.

The key here is to create simple, higher-level interfaces on top of Textile Threads that simplifies dealing with events and data, while still maintaining the power and flexibility of CQRS + ES. Developers should not have to learn a whole new set of terms and tools to take advantage of Textile's Threads capabilities. These simple, public-facing APIs will be comfortable to application developers looking to leverage highly-distributed systems that connect user-controlled data, with minimal configuration and maximum interoperability.

A database built on Threads would map database operations into Events. These Events would then mutate a View Model via its corresponding Reducer function, as per above. The actual View could be persisted or kept in-memory, and would be directly queried via a corresponding database-style APIs. Because Textile uses ES patterns, any delete-type operation would simply lead to an internal \emph{tombstone} Event, marking the database item for removal from the underlying materialized View.

\paragraph{Key/Value}

A key/value store built on Threads would ``map'' these database Events into key/value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del}. These Events would then mutate a map-like View Model. To complete the database-like API, the entire API could be encapsulated in a database sturcture that satisfies a given interface (see for example, Figure \ref{ex:KVStore}).

\begin{example}
\begin{minipage}{.45\textwidth}
\begin{lstlisting}
type TextileKVStore interface {
  Put(key string, value Node) error
  Get(key string) (Node, error)
  Del(key string) error
}
\end{lstlisting}
\end{minipage}
\caption{The Key-Value store interface.}
 \label{ex:KVStore}
\end{example}

\paragraph{No-SQL}

Other database abstractions include a no-sql style document store for storing and indexing arbitrary structs and/or JSON documents. The interface for such as store, again built using a materialized View, might look like Figure \ref{ex:DocStore}, where \texttt{Indexable} could be satisfied by any structure with a \texttt{Key} field and \texttt{Query} might be taken from the \texttt{go-datastore} interface library\footnote{\url{https://github.com/ipfs/go-datastore}} or similar.
